---
title: 'Reproducible Research: Peer Assessment 1'
author: "Jennifer Teed"
date: "May 17, 2015"
output:
  html_document:
    keep_md: yes
keep_md: yes
---

This assignment answers a number of questions about the Activity Monitoring Data dataset. 
Below I document the steps taken to answer the questions in the assignment.
Because I used RStudio, the knitr package was built into the environment.

```{r install, echo=FALSE, results='hide'}
#don't echo the code, and don't print out the results

```

## Loading and preprocessing the data


```{r}

# unzip the file
file <- unzip("activity.zip")
# set up the names and data types of the columns and then load the data
x<- read.csv("activity.csv", na.strings="NA", 
             col.names=c("steps", "date", "interval"), 
             colClasses=c("integer", "Date", "integer"))

len <- length(x$date)

```

The variables included in this dataset are:

  • steps: Number of steps taking in a 5-minute interval (with missing values are coded as NA )


  • date: The date on which the measurement was taken in YYYY-MM-DD format


  • interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of `r len` observations in this dataset.

## What is the mean of the total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

###1.Calculate the total number of steps taken per day.

```{r}

# Calculate the number of steps taken, per day, and put the answers into a vector.
v <- aggregate(x[, 1],list(x$date), sum, na.rm=TRUE)

# set the names of the columns here, to make the data more understandable
colnames(v) <- c("days", "total.steps")

```

###2.If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day.

```{r}

hist(v[ , 2], ylab="Frequency of occurance", xlab="Total Steps Reported in the Day", 
  main="Histogram of total number of steps taken each day")

```
From this data we can see that for about half of the days in the reporting period the person took between ten and fifteen thousand steps per day.


###3.Calculate and report the mean and median of the total number of steps taken per day

```{r}

# You could use summary here instead to get the answers, but then you can't plug the answers back into the 
# text below. Doing both here to double check answers
summary(v$total.steps)[c(3,4)]

meanVal <- mean(v[ , 2])
medianVal <- median(v[ , 2])

```

The mean total number steps taken per day is:  *`r meanVal`*.

The median total number steps taken per day is:  *`r medianVal`*.

The difference between my computed values and the values reported by summary are due to rounding and/or signifiant digits.


## What is the average daily activity pattern?

###1.Make a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
# save the intervals as a unique list, similar to a factor, for later use
# the format of the interval is HHMM, in 5 minute increments, so 100 is 1 hour, 0 minutes
interval.list <- unique(x$interval)

# group the data by interval, then for each interval find the mean number of steps, and plot the results. 

v2 <- aggregate(x$steps, list(x$interval), mean, na.rm=TRUE)
colnames(v2) <- c("interval", "ave.steps")

plot(interval.list,
     v2$ave.steps,
     type="l", 
     ylim= c(0, 3), 
     xlim = c(0,288), 
     ylab="average steps taken", 
     xlab="time of day (interval)",      
     main="Average Daily Activity")

```

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

## find the index of the max value of the vector above
index <-  which.max(v2$ave.steps)

## then find the index of that value to identify which interval it occurs in
interval <- interval.list[index]

```


The interval across all days which contains the maximum average number of steps is *`r interval`*.

The value of the average number of steps taken during that interval (across all days) is *`r index`*.


## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

###1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}

rows.missing.data<-sum(is.na(x))

```
There are `r rows.missing.data` rows with missing values in the dataset.

###2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Using the mean for the 5 minute interval is the method I am going to use to impute the missing data.


###3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
#start with the original data set, and copy it into a new dataset
nds <- x
#// interval.list, v2$ave.steps contain the values we want to swap
# create a boolean vector using which to find all the NAs (they're all in steps)

replaceNAs <- function(v1, v2)
{ 
    if(is.na(v1)) {v1 <- v2}
    else          {v1 <- v1}
}
  
nds$steps<- mapply(replaceNAs, nds$steps, v2$ave.steps) 

nds$steps

```

###4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
# Calculate the number of steps taken, per day, and put the answers into a vector.
v3 <- aggregate(nds[, 1],list(nds$date), sum, na.rm=TRUE)
colnames(v3) <- c("days", "total.steps")

hist(v3[ , 2], ylab="Frequency of occurance", xlab="Total Steps Reported in the Day", 
  main="Histogram of total number of steps taken each day")

# You could use summary here instead to get the answers, but then you can't plug the answers back into the 
# text below. Doing both here to double check answers
#summary(v3$total.steps)[c(3,4)]

meanVal <- mean(v3[ , 2])
medianVal <- median(v3[ , 2])

```

The mean total number steps taken per day is:  *`r meanVal`*.

The median total number steps taken per day is:  *`r medianVal`*.


```{r}

```

## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

###1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}


```

###2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
# what are the patterns on days that are M-F (compute)
# what are the patterns on other days
# plot the two, superimposed?
```


## End of document